Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-24 application.tex
re-written applications section

\section{Application to protoplanetary disks}
As an application of our results, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the simple criterion Eq. \ref{iso_vsi_cond}. The
condition derived below may be considered as a sufficient condition for the VSI.

@@

\subsection{Newtonian cooling}
Our aim is to extract from Eq. \ref{real_energy} the dimensionless
cooling time $\beta$ as used in our linear calculations. To do so, we
consider perturbations with small radial lengthscales $L\equiv
2\pi/k_x<l_\mathrm{rad}$ where 
\begin{align}
l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity. For
$L< l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the right hand side becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}
  t_c = \frac{c_v}{16\sigma_s\kappa_dT^3}.
\end{align}
For the fundamental isothermal VSI to operate, we need $\beta =
t_c\Omega_k< \beta_\mathrm{crit}$ as given by
Eq. \ref{iso_vsi_cond}. This translates to
\begin{align} 
  \frac{\mathcal{R}\Omega_k}{16\mu\sigma_s\kappa_dT^3} < \epsilon|q|. 
\end{align}

the Minimum Mass Solar Nebula (MMSN) described in \cite{chiang10}

>>>>

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do the thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental isothermal VSI, 
for which we can use the simple criterion Eq. \ref{iso_vsi_cond}. The
condition derived below may be considered as a sufficient condition for the VSI.

@@

Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
the thermal relaxation time $t_c$ as used in our linear
calculations. To do so, we first consider separately perturbations
with radial lengthscales $L\equiv 2\pi/k_x\ll l_\mathrm{rad}$ and 
$L\gg l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

\subsection{Newtonian cooling}
For $L\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

\subsection{Radiative diffusion}
For perturbation lengthscales $L\gg l_\mathrm{rad}$, let us assume that
temperature fluctuations are smoothed out by radiative diffusion. To
model this in linear theory, we choose the heating term such that
\begin{align}
  Q_+ - \nabla\cdot\bm{F} \equiv \nabla\cdot\left[k_\mathrm{rad}\nabla
    \left(T-T_{t=0}\right)\right], 
\end{align} 
where $k_\mathrm{rad}$ is the conduction coefficient to be defined
implicitly. Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} + \frac{P}{\rho T
    c_v}\nabla\cdot\left[k_\mathrm{rad}\nabla\left(T-T_{t=0}\right)\right].     
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}\label{diff_cool_proper}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right). 
\end{align}
This, of course, fundamentally changes the linear problem by
introducing higher order derivatives. We defer a proper analysis of
this problem to a follow-up study. Here, we are interested in
order-of-magnitude estimates. We thus proceed by assuming vertical
derivatives can be neglected in Eq. \ref{diff_cool_proper} and write
\begin{align}\label{diff_cool_approx}
  \frac{P}{\rho T c_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
    T\right) &\to -\frac{k_\mathrm{rad}}{\rho
    c_v}k_x^2P\left(\frac{\delta T}{T}\right)\notag\\
  &\equiv -\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right), 
\end{align}
%we are interested in radially-thin perturbations  
where $\eta=k_\mathrm{rad}/\rho c_v$ is the diffusion coefficient and
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 c_v}. 
\end{align}
From Eq. \ref{diff_cool_approx} we identify the thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}\equiv \frac{\Omega_k^{-1}}{\hat{\eta}\khat^2}, 
\end{align}
where $\hat{\eta} = \eta/H_\mathrm{iso}^2\Omega_k$ is the 
dimensionless diffusion coefficient defined for convenience, and is
approximately 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \simeq  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}

\subsection{Toy model for realistic thermal relaxation}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff} =
  \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
and Eq. \ref{eta_def}. Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$.

The dimensionless thermal relaxation time $\beta$ is then
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H_\mathrm{iso}^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right).
\end{align}

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).  
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model, 
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{9.4\times10^{-3}}{(\gamma
    -1)}\zeta^{2/7},
\end{align}
and we recall $\beta < \beta_\mathrm{crit}$ permits the fundamental 
isothermal VSI to operate.

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the MMSN with $\mu = 2$ and
$\gamma=1.4$. Although it is possible to satisfy
$\beta<\beta_\mathrm{crit}$ at small radii by choosing a sufficiently
large $\khat$, our linear calculations suggest growth rates eventually
decay as $\khat\to\infty$. We thus restrict our attention to
$\khat\lesssim 10^2$.

Fig. \ref{mmsn_bcrit_bcool} suggests that the fundamental isothermal
VSI can operate in the outer parts of the MMSN at radii on the order
of tens of AU. However, the thermal relaxation time $\beta$ is not
much shorter than $\beta_\mathrm{crit}$, suggesting the VSI may be
weak.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mmsn}  
  \caption{Dimensionless thermal relaxation timescales $\beta$ in the MMSN
    using Eq. \ref{beta_mmsn} for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot). The relaxation timescale
    $\beta_\mathrm{crit}$, below which the fundamental isothermal VSI
    operates, is shown as the solid line. 
    \label{mmsn_bcrit_bcool}}   
\end{figure}

\subsection{Numerical example}
Consider perturbations with $\khat=10$ in the 
the MMSN at $r=30\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2$. Then we find
the dimensionless thermal relaxation time, using
Eq. \ref{tc_newton_cool}, is
\begin{align*}
  \beta \simeq 0.046. 
\end{align*}
At $r=30\mathrm{AU}$, the disk aspect ratio is $\epsilon \simeq
0.058$ which implies, using Eq. \ref{iso_vsi_cond}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.062. 
\end{align*}
Thus, the disk supports the fundamental isothermal VSI, but only
marginally. Using the above value of $\beta$, MMSN disk parameters
and $\khat=10$ in our linear code, we find a growth rate of
\begin{align*}
  \nu = 2.7\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time of $t_g\simeq 60$
orbits.

 %only find one unstable mode -> close to marginal

 %larger kx
